### 2024-06-17 README.Rmd
More work

```{r include=FALSE}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(magick) # Advanced Graphics and Image-Processing in R
library(neuralnet) # Training of Neural Networks
```

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 70 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the importance of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

### Distract from learning

@@

In a sense, an AI is like a crutch. It can help someone to accomplish a task more easily, but it will not help if used constantly. However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles. Hence, the trunk will not slash a banana on its first attemt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature (in the sense that neural networks are blackboxes, and the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.], and not the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney. In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD). Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022. To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero money and still account for less than _one third of one percent point_ of Microsoft's revenue. A mid-sized Canadian university is like an ant compared to the economic boot of Microsoft. In truth, as the case of Solar Winds shows, Microsoft can refuse to fix a security flaw that leaves even the US Government vulnerable(for instance, to a Russian hack), and do so with little fear of ever facing accountability [@burke_microsoft_2024].] who fancy themselves creators but never learned (or forgot) how to code.).

@@

One could ask whether there is value in doing things more slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later, and perhaps even more efficiently.

### Learning the wrong stuff and deskilling

@@

Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly comes from using bigger datasets. There has been a rush for data^[This rush is perhaps more frenetic than gold rushes in the past. And, while the gold rushes of the 19th century were relatively democratic due to the low cost of immigration, the rush for data is largely dominated by a few, already ultra-wealthy players.], as developers try to improve their models by training them using more and more data. But, despite the large number of texts available in digital form, there is still only a finite amount of data to be had.

As of this writing, the big corporations that can afford to grab data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024].

IBM, for example, uses data drawn from the Project CodeNet Dataset. The data are "sourced" from two online judge web sites, AIZU Online Judge and AtCoder [@ibm_automatic_2021]. According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code." When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited". IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet. One might ask if IBM is doing all this research as a non-commercial effort? Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free, and use the systems for education and research only? What does "commercial uses" even mean?

As the sources of human-generated natural language data dry up, the few corporations developing AIs have turned to what might be the last juicy nuggets of text left on the internet.

OpenAI turned to Stack Overflow [@stack_overflow_stack_2024] for more data. Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world." The community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) with the way Stack Overflow dealt with OpenAI, and believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution. Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/). It is an open question how long Stack Overflow will remain a community-based space, when it treats the members of the community as unwitting providers of data for OpenAI.

Google, for its part, turned to Reddit for training data to the tune of \$60 million per year [@roth_google_2024]. As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures including a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].

@@

Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] (as of May, 2024) are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve AI? Probably not by much, since the key constraint is the amount of data available for training the deep neural networks behind GenAI.

Model poisoning.

Peak ChatGPT.

Develop a garbage skill

Words go here.

@@

But there is the following implication: once human-generated content is exhausted and/or has been overwhelmed by chatbot generated content, de-duplication becomes increasingly challenging.

The variance of the model's output is substantially reduced: the output will, of necessity, be less varied, less diverse, and more norm-conforming:
```{r data-vs-prediction-variance}
df_pred_all |> 
  group_by(type) |> 
  summarize(variance = var(y))
```

Tools to posion AI models:

@salman_raising_2023

@shan_nightshade_2024

@xu_shadowcast_2024

@@

@@

When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify. In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors^[A group that author Charles Stross colorfully [characterizes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html) as "fascist-adjacent straight white males with an unadmitted political agenda". This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay. On the other hand, their political agenda hides in plain sight. @thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible". Elon Musk, well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler, while defending the vandals who [terrorized Ottawa](https://fortune.com/2022/01/27/elon-musk-twitter-freedom-convoy-canadian-truckers-rule-covid-vaccine/) in 2022.] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity. Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in the future is well worth the sacrifices of a few billions of humans today. Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that those wealthy and well connected longtermists are willing to make.]).

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[this is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring. Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow, let alone in the face of powerful authoritarian regimes that might threaten their profits^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who should be concerned about maintaining an edge?]?

>>>>

```{r include=FALSE}
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(knitr)
library(magick) # Advanced Graphics and Image-Processing in R
library(neuralnet) # Training of Neural Networks
```

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 75 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

### Distract from learning and deskilling

@@

In a sense, an AI is like a crutch. It can help someone to accomplish a task more easily, but it will not help if used constantly. However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle that the end user cannot hope to master^[Reports of people getting their fingers crushed by Tesla's cybertruck could be a result of 1) people trusting the technology more than they should; and 2) not understanding the technology: apparently the algorithm uses increasing force to close the trunk if it fails to latch on back-to-back cycles. Hence, the trunk will not slash a banana on its first attempt to latch, but will crush a finger on the second attempt [@lambert_tesla_2024].] due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox where things change based on the interests of their corporate developers^[Sometimes said interest may be relatively benign, like avoiding reputational damage [by correcting a wrong](https://www.bbc.com/news/technology-65202597), but there is no guarantee that corporate interests will always align with those of the public, and in fact, they seldom will as long as there are only a handful of American companies that offer an "essential" service.] instead of the interests of their relatively small customers^[And this refers not only to individuals whose skills are yoked to ChatGPT and MidJourney. In 2022, McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD). Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022. To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue. A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft. In truth, Microsoft feels that it can stand up to even the US Government. As seen in the case of Solar Winds, Microsoft refused to fix a known security flaw that left the US Government vulnerable to a Russian hack, and did so with little fear of ever facing accountability [@burke_microsoft_2024].].

@@

One could ask whether there is not more value in doing things slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later--perhaps even more efficiently.

@@

Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly are the result of using bigger datasets. This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past. And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by Venture Capital (VC) and large corporations.]. But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.

As of this writing, the few big players that can afford to grab the data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024]. IBM, for example, uses data drawn from the Project CodeNet Dataset. The data are "sourced" from two online judge web sites, AIZU Online Judge and AtCoder [@ibm_automatic_2021]. According to IBM "[the] end goal of CodeNet is to enable developers to create systems that can modernize existing codebases, as well as fix errors and security vulnerabilities in code". When we inspect the Terms of Use of AIZU (a Japanese academic project), "Commercial uses of the registered codes are prohibited". IBM studiously avoids the word "commercial" in its posts about LLM, foundational models, and CodeNet. One might ask if IBM is doing all this research as a non-commercial effort? Will those developers creating systems, modernizing databases, fixing errors, etc., be able to access this for free, and use the systems for education and research only? What does "commercial uses" even mean?

As the fountain of human-generated content, generously shared by countless people over a period of decades, becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.

OpenAI, as a leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024]. Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world". Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not_ do. Some members of the community have protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/). It is an open question how long Stack Overflow will remain a community-based space, after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.

Google, for its part, turned to Reddit for training data to the tune of \$60 million per year [@roth_google_2024]. As Reddit content begins to creep into Google's AI-assisted search, the results of this deal have turned grotesque, with reports of AI Overview failures that include a suggestion to put non-toxic glue to pizza sauce to keep the cheese from sliding [@koebler__google_2024]^[The headline of the 404 Media report was "Google Is Paying Reddit $60 Million for Fucksmith to Tell Its Users to Eat Glue".].

@@

Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve GenAIs? Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind GenAI [@villalobos_position_2024]. And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[Disturbing as this is, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information. It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.]. As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites^[A blog post on a website called _Ancient Rome_ by a certain Ellen Hunter (whose profile picture is likely the output of a generative AI and who does not seem to exist elsewhere) discusses the topic of windows in classical Rome and Greece [@hunter_did_2023]. Under subheading 4.1 (_What Existed Before Windows?_), the blog explains that "MS-DOS was a disk operating system that was developed by Microsoft. It was the predecessor to the Windows operating system and was used on IBM PC compatible computers. MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s". Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update" [e.g., @swain_deep_2023; @al-worafi_pharmacy_2023; @sumit_dynamic_2024; @zhang_retracted_2024].].

What can we expect as, in what probably is inevitable by now, future GenAI models begin to be trained using data that was generated by previous GenAIs? Revisiting the examples in a preceding section ([But really, what do AIs do?](#really-what-do-they-do)) we can see that the predictions of the models are more uniform than the original data used for training. Technically, the variance of a model's output is substantially reduced compared to the variance of the training data: the outputs of a model will necessarily (as they regress towards the mean) be less varied, less diverse, and more norm-conforming (see Table 1).

```{r data-vs-prediction-variance}
df_pred_all |> 
  group_by(type) |> 
  summarize(variance = var(y)) |>
  transmute(Model = factor(type, 
                       levels = c("training data", "regression output", "nn1 output", "nn2 output"),
                       labels = c("Training data", "Linear regression", "Neural network 1", "Neural network 2"),
                       ordered = TRUE),
         Variance = round(variance, 2)) |>
  kable(caption = "\\ref{tab:data-vs-prediction-variance}Table 1. Variance reduction in model outputs")
```

The implication should be clear: training models with the output of models is a recipe for less varied outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to a bland version of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways, too subtle for the changes to be detectable by a human, but deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

All in all, the evidence suggests that, as fresh data for training future GenAIs dries up, the performance of these machines may not improve drastically from what we have seen so far, that is, flashy results that bear no relationship to the truth, being in fact amoral bullshit-producing machines^[American philosopher Harry G. Frankfurt defines _bullshit_ as speech intended to persuade without regards to the truth [@frankfurt_bullshit_2005]. AIs, lacking intent, do not lie--but they do bullshit.]. In fact, the performance may begin to deteriorate as the output of GenAIs begins to be used to train other GenAIs. This effect could easily be compounded if, as is likely the case, human-generated content is exhausted, and/or is overwhelmed by GenAI content, and/or is corrupted by purposefully bad data supplied by actors ranging from outraged creators to adversarial regimes. All of which begs the following questions. How useful will AIs be in the future? How persistent will skills like prompt "engineering" be? Will learning to prompt an AI be worth sacrificing the opportunity to develop non-AI-powered skills like coding, writing, and thinking? Who, after all, stands to benefit the most by this form of deskilling, in exchange for working to develop an ability that may turn out to be of ephemeral value at best and a dead-end at worst?

<!--
Develop a garbage skill

@@

But there is the following implication: , de-duplication becomes increasingly challenging.
-->

@@

Deskilling leads to...

@@

When using a Generative AI we must ask whose perspectives the tool will tend to reproduce and amplify. In this sense, numerous observers [e.g., @ahuja_we_2023; @stross_tech_2023; @gebru_tescreal_2024] have noted the ideological leanings of various actors in the AI scene, an elite world of tech investors^[A group that author Charles Stross colorfully [characterizes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html) as "fascist-adjacent straight white males with an unadmitted political agenda". This characterization is somewhat unfair: both [Sam Altman](https://www.advocate.com/news/sam-altman-openai-exclusive) and [Peter Thiel](https://www.npr.org/2016/07/21/486966882/i-am-proud-to-be-gay-tech-investor-peter-thiel-tells-gop-convention) are openly gay. On the other hand, their political agenda hides in plain sight. @thiel_education_2009 has written about how he "no longer believe[s] that freedom and democracy are compatible". Sam Bankman-Fried's FTX donated millions to a group with racist ties [@wilson_sam_2024]. And Elon Musk? Well, Elon Musk once shared a meme comparing Canada's PM Justin Trudeau to Hitler while defending the vandals who terrorized Ottawa in 2022 [@henderson_and_elon_2022].] known for fringe ideologies that combine a peculiar mix of uber-techno-optimism and existential fright (including transhumanism, extropianism, singularitarianism, and cosmism)^[A quick summary of these beliefs is that humans (or at lest some of them, those who can afford to do so) are destined to use technology to indefinitely extend life and conquer the stars...but maybe not before super-intelligent AIs destroy humanity. Just in case, wealthy tech barons like Peter Thiel and Sam Altman, have made suitable [escape plans](https://www.theguardian.com/news/2018/feb/15/why-silicon-valley-billionaires-are-prepping-for-the-apocalypse-in-new-zealand).], paired with extreme theories of justice (including rationalism, effective altruism^[A movement exposed to the harsh light of public opinion for being the philosophy espoused by convicted fraudster and money launderer Sam Bankman-Fried [@crook_effective_2023].], and longtermism^[A drastic form of utilitarianism convinced that the well-being of trillions of humans in an imagined distant, star-faring future, is well worth the sacrifices of a few billions of humans today. Curiously, the contribution of the titans of technology promoting AI is not to sacrifice themselves: as evidence of the abuses inflicted in the name of innovation continues to pile up, it is clear that some of us will suffer and die--but that is a sacrifice that wealthy and well connected longtermists are willing to make.]).

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring. Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who should be concerned about maintaining an edge?]?
